


\chapter{
 2012: A Spatial Capture-Recapture Odyssey
 }

\markboth{The End}{}
\label{chapt.final}

\vspace{0.3cm}


\vspace{2in}

Capture recapture methods have been a cornerstone of ecological
modeling and analysis for decades.  Yet there are essentially no real
capture-recapture data sets that come {\it without} auxiliary spatial
information about location of capture (but sometimes such information
is thrown into the trashcan).  As such, classical capture-recapture
models are usually not the right tool for the job of analyzing real
data sets, unless you happen to study fish populations in tanks. 
       % XXXX I would tone this down and say that "virtually all
           % classical analyses can be improved by making use of the
           % available spatial data.
Instead, biologists need methods -- spatial
capture-recapture methods -- that make use of spatial
information in their capture-recapture data sets.

Spatial capture-recapture methods resolve the essential problem of
capture-recapture, that of estimating population size,
% XXXX I would add: "and density,"
but, by
explicitly linking space occupied by individuals with the spatial
location or region of sampling, they do so in a manner that provides a
holistic framework for answering ecological questions related to the
spatial structure of populations -- movement, space usage, spatial
variation in density, landscape connectivity, and other things
(probably).  Thus, spatial capture-recapture methods enable ecologists
to integrate science with their capture-recapture models based on
individual level encounter data.

We provided an overview and synthesis of what is currently known (or,
rather, published in the literature),
around mid-2012, related to spatial capture-recapture methods. There
are many emerging topics which we have not covered either because lack
of technical knowledge, or satisfactory development, or no good
framework for implementation......

\section{Emerging Topics}

A number of topis of importance, or that people are working on, or
that might make good PhD or Masters prodjects.


\subsection{Inhomogeneous Point Processes}

In currently developing work, we propose a model that accounts for
spatial variation in home range density and potential interactions
between individuals' home ranges.  This model lets the activity
centers follow an inhomogeneous Strauss process
\citep{strauss:1975,handbook:2010}, which allows for spatial variation
in the home range intensity, and includes a parameter that determines
the strength of repulsion between home ranges.  We show via a
simulation study that properly accounting for interactions between
individuals can provide a substantial improvement in estimating
population size.  For simulated data generated with interaction, the
usual independence model has a significant bias for the population
size, and generally has larger uncertainty for the population size
than the proposed Strauss process model.

While the Strauss model is intuitive and shows great potential, it
presents computational challenges.  First, the likelihood includes a
high-dimensional integral that has no closed form.  Extending related
work for categorical Markov random fields
\citep{green:2002,smith:2006}, we develop an approximation to the
Strauss likelihood which allows for posterior sampling.  Second, in
our Bayesian analysis, the population size is treated as an unknown
parameter to be updated using the data.  As the population size
varies, so does the dimension of the likelihood, and thus the
posterior.  We overcome this dimension-changing problem using an
auxiliary variable scheme in the Markov chain Monte Carlo algorithm.
Our analysis of simulated data verifies that this computational
approach leads to reliable inference.


\subsection{Combining data from different surveys}

In some instances, researchers apply different survey techniques to
the population of interest, because they yield complementary
information. For example, Camera trapping is the prime tool for
estimating population size/density and other demographic parameters
for uniquely marked species; genetic surveys can yield additional
information on the genetic diversity and health of a population that
we cannot study using camera traps. At the same time, genetic surveys,
when samples are analyszed to the individual level, also yield spatial
capture recapture data (see Chapt. \ref{chapt.search-encounter}). In
this situation, we have two data sets at hand that carry information
on animal density, and we should be able to get more precise estimates
of density if we combine these two data sets into a single SCR model.

\citet{gopalaswamy_etal:2012} developed two approaches to combining
data from different survey types. In the first case, both surveys are
carried out at the same time, so that we can assume that they both
sample the same -- closed -- animal population, i.e., there are no
possible changes in population density between the two surveys. For
camera trapping and genetic surveys, we cannot match records of
individuals between the two data sets. As a consequence, in the
combined model there are two sets of $z_i$, say, $z^{C}_{i}$ for the
camera trap data and $z^{G}_{i}$ for the genetic survey data. But
defining a single state-space around both sets of survey locations, we
can define a single data augmentation paramter, $\psi$, that refers to
both these sets of $mathbf{z}$.  {\bf XXX ANDY; WE STILL HAVE TWO ESTIMATES
OF N AND THUS D; HOW DID YOU CHOOSE? OR DID YOU AVERAGE? I DONT FIND
ANY INFORMATION ON THAT IN ARJUNS MS XXXXX}

Further, since the scale paramter of the trap encounter model (in
\citet{gopalaswamy_etal:2012} the Gaussian model}), $\sigma$, is
related to animal movement, we can expect this parameter to be the
same for both surveys and share it across both data sets. Finally, we
need to define separate baseline detection parameters, say
$\lambda_{0}^{C}$ and $\lambda_{0}^{G}$, because the processes
governing detection by the survey methods should be
different. \citet{gopalaswamy_etal:2012} found that this combined
model did indeed produce a more precise density estimate, compared to
single-data set models.  We can, of course, imagine other
parameterizations for this combined model -- we could specify both
$\psi$ or $\sigma$ as survey specific, if we have reason to believe
they changed between surveys, and we refer you to
\citet{gopalaswamy_etal:2012} for more details on these alternative
parameterizations.

A second approach of using information from one survey in the analysis
of a second survey (that maybe does not yield quite as much data as
the primary survey) is by analyzing your primary data set alone, then
taking the posterior distribution of a parameter both surveys share
and using it as an informative prior distribution in the analysis of
the second data set. \citet{gopalaswamy_etal:2012} refer to this as
the stepwise approach, and they implemented this approach by equating
the mean and variance of the posterior distribution of $\psi$ and
$\sigma$ from the photographic survey to the mean and variance of a
beta and a gamma prior for these paramters, respectively, for the
genetic survey. The authors found that this approach produced almost
identical density estimates compared to the combined model approach
descibed above.

In summary, no matter which approach is chosen, combining data across
surveys can help researchers to obtain more precise population
estimates, which is especially valuable when dealing with rare and
elusive species like big cats that almost always will produce sparse
individual data sets. Some thought has to go into the assumptions
underlying combining data -- for example, if there is a chance that
population size may have changed between surveys, it might not be
approapriate to have $\psi$ be a parameter shared across surveys; but
the assumption that $\sigma$ remains reasonably constant might still
be valid. The paper by \citet{gopalaswamy_etal:2012} considers the
situation where we have two SCR data sets, but we can imagine
combining SCR data with other sources of information, such as
telemetry data (see Chapt. \ref{chapt.partialID} and
Chapt. \ref{chapt.rsf} for examples), and possibly opportunistic
observations, although to our knowledge this latter issue has not been
tackeld in the context of SCR, yet.


\subsection{Imperfect identification of individuals}

Imperfect identification of individuals can happen in a variety of
ways. In genetic surveys there is usually some probability of
mis-identification of individuals due to genotyping error
(e.g. \citet{lukacs_burnham:2005}). In camera trap survey a different
type of imperfect identification can occur when only the only one
flank of an animal is recorded in a detection event and that imagine
cannot be matched to any of the individuals identified by both
flanks. If that case, we can match single-flank pictures with the same
side flank pictures, but not with opposite side flank pictures and
thus cannot construct definite encounter histories for these
single-flank individuals (a right flank and a left flank picture could
be the same individual, or could be from two distinct
individuals. Finally, in Chapt. \ref{chapt.partialID}, in the context
of mark-resight models, we discussed the case where individuals can
either not definitely be identified as marked or not -- a violation of
a basic mark-resight assumption, and developed an approach to dealing
with the situation where we can always tell if an animal is marked or
not, but we are not always able to ascertain its individual identity.

In non-spatial capture recapture some efforts have been made to
formally deal with misidentification. \citet{stevick_etal:2001}
address this problem by double-sampling to derive an error rate for
genetic identification, and then including this error rate as a knwon
constant into a Lincoln-Petersen estimator of
abundance. \citet{lukacs_burnham:2005} develop an approach that
includes an additional parameter in the model -- the probability of a
genotype being identified correctly, which is estimated as part of the
model likelihood. \citet{link_etal:2010} developed an approach towards
solving the same problem implemented in a Bayesian framework that
relaxes some of the assumptions of the initial approach.
\citet{yoshizaki_etal:2009} deal with misidentification from camera
trap pictures due to evolving markes (i.e., natural marks that change
over time, such as scars). This situation is different from the
genotyping error one. Here, a change in marks creates a supposedly
`new' individual that can be recaptured several times, while the
original individual is never captured again (its mark is no longer in
the population). In contrast, in genotyping error it is assumed that
misidentification creates a `new' individual that is never observed
again, because each error leads to a new unique
genotype. \citet{yoshizaki_etal:2009} approach this situation
similarly, by including a parameter describing the probability of
correctly identifying an individual upon recapture (the parameter can
also be interpreted as the probability that a mark does not change
between capture occasions). Because of the dependencies between true
and false detection histories (when a `new' individual is created, the
`real' one can no longer be recaptured), the standard multinomial
approach to coming up with a model likelihood does not work and
implementing the model in a maximum likelihood framework is
difficult. The authors instead demostrate an implementation of the
model based on minimizing a function of the squared differences
between the observed and expected frequencies of the observed capture
histories.

To our knowledge no attempts have been made to deal with
misidentification in an SCR framework. While all of the mis-ID cases
described above require distinct approaches, we believe that there is
one unifying theme to all of them: the location of the un or
potentially mis-identified records and the resulting probabilities of
belonging to certain individuals conditional on their activity
centers. For example, a right flank and a left flank camera trap
picture that are taken at two neighboring camera traps are more likely
to belong to the same individual that a right and a left flank picture
taken at cameras at opposing ends of the trap array, especially if
animal movement is smaller than the extent of the trap array; an
unidentified record of a marked individual in a mark-resight survey is
more likely to belong to a marked individual whose activity center is
close by, than to an individual whose activity center is located far
away; and so forth. Formally developing misidentification models
should be a focus for future SCR model development.

\subsection{Three dimensional space}

Throughout this book we have treated space as
two-dimensional, meaning that activity centers are assumed to occur on
the real plane. This approximation of reality is reasonable for many
terrestrial species, but aquatic organisms, especially marine animals
move about in three-dimensional space. Treating space as
three-dimensional could also conceivably be useful in studies of
flying organisms, aquatic organisms,
or species that use multiple strata of tall forests; however, we
suspect that two dimensional models of space should suffice in such
contexts. Regardless, a three-dimensional view of space requires that
activity centers $\bf s_i$ are indexed by
$x,y,z$ coordinates. In theory, this presents no problem whatsoever. In
practice, estimation based on integrated likelihood methods must
involve a three-dimensional integration. This will clearly be more
computationally demanding, but it should be possible using packages
such as {\tt R2Cuba}.


\subsection{Gregarious species}

Many social species move about in large groups rather than as single
individuals. Even species regarded as solitary often join family
groups for some portion of their life cycle. The consequences of
gregariousness?? are x-fold....

To account for this, we change our definition of $s_i$ from the
location of an individual's activity center, to the location of a
group's activity center. We then expand our model to include a
submodel for group size, and we can estimate both the density of group
activity centers and total population size.

\subsection{Design}


\subsection{Borchers acoustic}

from ISEC

\subsection{Single Catch Traps}


\subsection{Model Fit and Selection}


Calibration of GoF under meaningful alternatives

Calibration of AIC/DIC



\subsection{Dynamics}

Dynamic point processes:
Movement, Dispersal, Migration
Transience.


\subsection{Other things}


Models for unmarked or partially marked individuals integrated with RSF data from telemetry

Occupancy and counts data + SCR data (AOAS and Sollmann et al.)

Spatial genetics  -- can use SCR to study gene flow, related things....

SCR on dendritic networks (streams and trails).

Effect on design of telemetry and other things



\section{Grand Summary}


Historically the main use of capture-recapture was to get population
size estimates, but
SCR models allow you to address basic and applied questions of
population ecology from individual encounter history data. Problems
having to do with movement, space usage, landscape connectivity,
etc... This is a BFD.













