Resource Selection and Space Usage


Most SCR approaches use a symmetric and stationary encounter probability 
model which implies that space usage is a decreasing function of distance
from an individual's home range center.  This is often not very realistic for most
applications.  In this chapter, we extend SCR models to
incorporate models of resource selection, such as when one or more
explicit landscape covariates are available which the investigator
believes might affect how individual animals use space within their
home range.  We present a method that integrates a
standard family of resource selection models based on auxiliary
telemetry data into the capture-recapture model for encounter
probability.   The important distinction between SCR and resource
selection function (RSF)
studies is that, in SCR studies, encounter of individuals is imperfect
(i.e., ``$p<1$'') whereas, with RSF data obtained by telemetry,
encounter is perfect.  Thus in this chapter, we argue that SCR and telemetry data 
can therefore be combined in the same likelihood by formally
recognizing this distinction in the model.  To do so, we first identify possible
models of space usage, including those commonly used in the resource
selection literature, and provide an example using simulated data.  Then we
demonstrate a model for integrating capture-recapture data with RSF data under
a Poisson model of space usage.  We present and analyze
data from a study of black bears in southwest NY, USA.  We fit a variety of models described
in the chapter to the data to highlight the differences of including only SCR data or SCR plus RSF
 data.  We also provide a simulation study to validate the model and for the reader to become
more familar with the models and data. 
This integration of SCR and RSF models wil allow researchers to model how the
landscape and habitat influence the movement and space use of
individuals around their home range, using non-invasively collected
capture-recapture data that can be augmented with telemetry data.

Key Words:  use decisions, telemetry, RSF, black bears, habitat selection, integrated data sources,
landscape