
The project owes a great intellectual debt to Jim Nichols, who has
been an extraordinary mentor and colleague,
and who generously shared his
astounding insight into animal sampling and modeling problems, his
knowledge
of the literature and history of abundance and density estimation
and, most importantly, his extremely valuable time.
He has
been
an 
extremely helpful guy on all fronts. We are honored that Jim
agreed to write the Foreword to introduce the book. 
 We thank Marc K\'{e}ry
for being a great friend and colleague, and for his creativity, 
energy, and enthusiasm in developing new ideas and
presenting workshops on hierarchical modeling in ecology.

{\bf Special thanks} to the following people: (1) Kimberly Gazenski whose support was
invaluable. She worked on administrative, technical and editorial
aspects of the whole project. She maintained the \textsc{Bib}\TeX~database,
worked on the GitHub repository that housed the \LaTeX~and \R~package
source trees, edited \LaTeX~files, tested R scripts, did GIS and
\R~programming, analysis, debugging, and graphics;  (2) Our WCS Tiger
program colleagues K. Ullas Karanth and Arjun Gopalaswamy for
continued support and collaboration on SCR problems;  (3) Sarah
Converse, our PWRC colleague, for her interest in SCR models and
developing a number of methodological and application papers related
to multi-session models, providing feedback on draft material, and
friendship;  (4) Murray Efford whose seminal 2004 Oikos paper first introduced spatial
capture-recapture models.  His \R~package \mbox{\tt secr} 
is a powerful tool for analyzing
spatial capture recapture data used
throughout the book. Murray also answered many questions regarding \mbox{\tt secr} that 
were helpful in developing our applications and examples.

We thank the following people for providing data, photographs or figures:
Michael Wegan, Paul Curtis, and Raymond Rainbolt (black bear data from Fort Drum, NY);
Audrey Magoun (wolverine data); Cat Sun and Angela Fuller (black bear
data in Chapt. 13); Joshua Raabe and Joseph Hightower (American shad
photo and data in Chap. \ref{chapt.open}); Erin Zylstra (tortoise
data in Chapt. \ref{chapt.closed}); Martha (Liz) Rutledge (Canada geese
data and picture in \ref{chapt.partialID});  Craig Thompson for the
fisher data (Chapt. \ref{chapt.search-encounter});
Jerrold Belant (black bear data
in Chapt. \ref{chapt.design}); Theodore Simons, Allan O'Connell,
Arielle Parsons, and Jessica Stocking (raccoon data in Chapt.
\ref{chapt.partialID}); Agustin Paviolo (jaguar data in Chapt. 11, and
the cover image).

We thank the following people for reviewing one or more draft chapters
and giving feedback along the way:
David Borchers,
Bob Dorazio,
Tabitha Graves,
Marc K\'{e}ry,   %do we need him twice?
Brett McClintock,
Allan O'Connell,
Krishna Pacifici,
Agustin Paviolo,
Brian Reich,
Robin Russell,
Sabrina Servanty,
Cat Sun,
and
Chris Sutherland.

%Additionally, many others provided us with encouragement and feedback throughout this project and for that 
%we thank our colleagues, friends, and families for their continued support.

Additionally, many colleagues and friends, as well as our families, provided us with encouragement and 
feedback throughout this project and
we thank them for their continued support.